---
title: "Week 3 - In Class"
author: "Vicki Yiannakis"
format: pdf
geometry:
      - top=30mm
      - left=20mm
      - heightrounded
---
```{r} 
#| output: false
#| include: false
#| warning: false
library(tidyverse)
library(knitr)
library(kableExtra)
install.packages("countrycode")
library(countrycode)
install.packages("janitor")
library(janitor)
```


First we will read in the maternal_mortality CSV
```{r}
#| echo: TRUE
matmor0 <- read.csv("data/raw/maternal_mortality.csv", header = TRUE)
glimpse(matmor0)
```

We want to create a function that uses the below code pivots this to long form 
```{r}
# Change wide to long format
matmor_long <- matmor0 |>
  pivot_longer(cols = starts_with("X"), #selecting all cols that start with x 
                 names_to = "year", #creating new variable called Year 
                 names_prefix = "X", # removes X from year column (so only numerical shows)
                 values_to = "maternal_mortality") |> #values of maternal mortality 
  mutate(year = as.numeric(year)) |> # change year to numeric As it was stored as a character var
  select(iso, year, maternal_mortality) #output data 
```


NOTES:
names_prefix = A regular expression used to remove matching text from the start of each variable name.

To create a function, we need to figure out our arguments 
```{r}
long_form <- function(x, varname) {
    x|>
    pivot_longer(cols = starts_with("X"), #selecting all cols that start with x 
                 names_to = "year", #creating new variable called Year 
                 names_prefix = "X", # removes X from year column (so only numerical shows)
                 values_to = varname)|>
        mutate(year = as.numeric(year)) |>
        select(iso,year, all_of(varname)) #values of maternal mortality 
} 

#now we can try this in the original mortality data
long_maternal <- long_form(matmor0, "maternal_mortaltiy")

```

Now we will try with the other datasets! 
```{r}

neonatal <- read.csv("data/raw/neonatal_mortality.csv", header = TRUE)
neonatal_long <- long_form(neonatal, "Neonatal_Mortality")


infant <- read.csv("data/raw/infant_mortality.csv", header=TRUE)
infant_long <- long_form(infant, "Infant_Mortality")

under5 <- read.csv("data/raw/under5_mortality.csv", header=TRUE)
under5_long <- long_form(under5, "Under5_Mortality")
```

2. Prepare disaster data

```{r}
disaster <- read.csv("data/raw/disaster.csv", header=TRUE)
disaster1 <- clean_names(disaster)
```

Now we will filter to only include the variables we need
Make sure to check variable types! 

```{r}
class(disaster1$year)
class(disaster1$disaster_type)

```

```{r|}

disaster_filter <- disaster1 %>% filter(year >= 2009 & year <= 2019,
                                        disaster_type == "Earthquake" | 
                                            disaster_type == "Drought") %>% 
                    select(year, iso, disaster_type)

kable(disaster_filter|> slice(1:10))                                
```

Create a dummy variable! 
```{r}
disaster_filter2 <- disaster_filter %>% 
    mutate(earthquake = if_else(disaster_type=="Earthquake", 1, 0),
           drought = if_else(disaster_type=="Drought", 1, 0)) %>%
                    mutate((across(where(is.numeric), as.integer)))
                                                

kable(disaster_filter2|> slice(1:10))
```

3. Prepare the disaster data file 

We are creating a binary variable to ascertain whether they had armed_conflict in that year 

```{r}
conflict <- read.csv("data/raw/conflict.csv", header=TRUE)

#we will need to create a binary_conflict variable for each year 1999 to 2018 (19 total)
#first we need to filter and sum up the number of conflicts each year

conflict1 <- conflict %>% group_by(iso,year)%>% summarise(total_conflicts=sum(best))

#now we are going to create a binary variable in which year is lagged 


conflict1 <- conflict %>% group_by(iso,year)%>% summarise(total_conflicts=sum(best, na.rm=TRUE))%>% mutate(binary_conflict = if_else(lead(total_conflicts)>0, 1, 0))
```


**MERGING ALL DATASETS** 
We have:
- conflict1 
- disaster_filter2
- infant_long
- matmor_long
- neonatal_long
- under5_long
They all have the ISO term. Hence they will be merged this way 

```{r}
full_dataset_list <- list(matmor_long, neonatal_long, infant_long, under5_long, disaster_filter2, conflict1)

merged_all <- full_dataset_list %>% reduce(full_join, by = c("iso", "year"))

#save as a CSV file and add to proccessed folder 
write.csv(merged_all, "Merged_Data.csv", row.names = FALSE)


